Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-09-29 psy_peds_3.Rmd
new version

listOfPackages <- c("tidyverse", "RColorBrewer", "knitr", "kableExtra", "tsModel")
paket(listOfPackages)
```

@@

### Extract the patient summary and observation information. 
demo_raw <- files[["patientSummary"]]
obs_raw <- files[["patientObservations"]]
clinical_raw <- files[["patientClinicalCourse"]]

@@

```{r}
# Adding filtering on in_hospital == 1 here to reduce subsequent compute time
clinical_raw <- filter(clinical_raw, in_hospital == 1)
```

@@

```{r}
#icdCodes <- read.csv("public-data/pediatric_psychiatric_ICD10_codes.csv", header = TRUE, colClasses = "character") %>% 
#  rename(disorder_group = Mental.Health.Disorder.Group, 
#         description = Description, 
#         originalCode = ICD10_Code) %>%
#  mutate( ICD10_Code= ifelse( nchar( originalCode) > 3, paste0(substr( originalCode, 1, 3), ".", substr(  originalCode, 4, nchar(originalCode))),originalCode)) %>%
#  select( disorder_group, ICD10_Code, description)

icdCodes <- read.csv("public-data/pediatric_psychiatric_ICD10_codesV2.csv", header = TRUE, colClasses = "character") %>%
   filter( flag %in% c("originalCodes", "FrenchSpecificCodes")) %>%
   select( disorder_group, ICD10_Code, description )
```

@@

clinical_raw_filter <- clinical_raw %>%
  filter( len_hospitalisation > min_hosp_days, 
          cohort  %in% c("AllAdm")) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic", "during_pandemic"))
```

```{r}
stopifnot(unique(clinical_raw_filter$cohort) == "AllAdm")
all_codes <- clinical_raw_filter %>% 
  left_join( obs_raw,  by = c("patient_num", "days_since_admission")) %>%
  filter( concept_type == 'DIAG-ICD10')
stopifnot(unique(all_codes$in_hospital) == 1)
stopifnot(unique(all_codes$len_hospitalisation > min_hosp_days))
```

@@

### New hospitalizations
```{r, fig.width=12, fig.height=12}
new_hospitalizations_per_day <- all_codes %>%
  group_by(patient_num, n_hospitalisation) %>% slice(which.min(date)) %>%
  filter( in_hospital == 1, 
         date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date, weeks, month) %>%
  summarise(count = n_distinct(patient_num)) %>% 
  ungroup()

@@

nh_day <- ggplot(new_hospitalizations_per_day,
                 aes(x = date, y = count))  +
  geom_point() +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(x = "Calendar date",
       y = "Patients count", 
       title = "New admissions (per day)")

nh_week <- ggplot(new_hospitalizations_per_week, aes(x = weeks,
                       y = count))  +
  geom_point() +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(x = "Calendar date",
       y = "Patients count", 
       title = "New admissions (per week)")

nh_month <- ggplot(new_hospitalizations_per_month, aes(x = month,
                       y = count))  +
  geom_point() +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(x = "Calendar date",
       y = "Patients count", 
       title = "New admissions (per month)")

require(gridExtra)
grid.arrange(nh_day,
             nh_week,
             nh_month,
             ncol=1)
```

### Total hospitalize patients
```{r, fig.height=12, fig.width=12}
total_hospitalizations_per_day <- all_codes %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count = n_distinct(patient_num))

total_hospitalizations_per_week <- all_codes %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks ) %>%
  summarise(count = n_distinct(patient_num))

total_hospitalizations_per_month <- all_codes %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month ) %>%
  summarise(count = n_distinct(patient_num))

@@

th_week <- ggplot(total_hospitalizations_per_week, aes(x = weeks,
                       y = count))  +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(x = "Calendar date",
       y = "Patients count", 
       title = "Total admissions (per week)")

th_month <- ggplot(total_hospitalizations_per_month, aes(x = month,
                       y = count))  +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(x = "Calendar date",
       y = "Patients count", 
       title = "Currently hospitalized patients (per month)")

@@

### Calculate how many of the mental codes are in the dataset
mentalHealthCodes <- unique( all_codes %>%
  dplyr::filter( concept_type == "DIAG-ICD10", 
                 concept_code %in% icdCodes$ICD10_Code ) %>%
  dplyr::mutate( concept_code = as.character( concept_code ), 
                 concept_length = nchar( concept_code ), 
                 concept_check = ifelse( concept_length == 4 | concept_length > 8, "Check", "OK")) %>%
  dplyr::select( concept_code, concept_length, concept_check) )

### Add the mental ICD code description and filter 
```{r message=FALSE, warning=FALSE}
# Getting site name, will be used later in the analysis
site <- unique(as.character(demo_raw$siteid))

@@

all_codes_desc <- left_join( all_codes_desc, icdCategory) %>%
  mutate( icd_description_category = replace_na( ifelse( concept_code %in% icdCodes$ICD10_Code, "Psy", icd_description_category), "Others"))

input_data <- all_codes_desc %>% 
  filter(concept_code %in% icdCodes$ICD10_Code)

if( raceAvailable == TRUE ){
  input_data <- input_data %>%
    left_join( race_raw, by = c("patient_num", "cohort") )
}
```

@@

```{r}
total_hospitalizations_per_day_psy <- input_data %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count_psy = n_distinct(patient_num)) %>%
  left_join( total_hospitalizations_per_day ) %>%
  mutate( period = ifelse( date < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic"), 
                percentage = count_psy * 100 / count)

thpd_psyPer <- total_hospitalizations_per_day_psy %>%
  ggplot(aes(x = date, y = percentage, fill = period)) +
  #geom_line()+
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  labs(y = "Percentage",
       x = "Date (by day)",
       title = "Percentage of patients with mental health related ICD codes (daily)") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyPer

thpd_psyCount <- total_hospitalizations_per_day_psy %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = date, y = value, fill = counts, color = counts)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Date (by day)",
       title = "Daily patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

>>>>

listOfPackages <- c("tidyverse", "RColorBrewer", "knitr", "kableExtra", "tsModel", "gridExtra")
paket(listOfPackages)
```

@@

### Extract the patient summary and observation information. 
demo_raw <- files[["patientSummary"]] %>% filter(cohort == "AllAdm")
obs_raw <- files[["patientObservations"]] %>% filter(cohort == "AllAdm")
clinical_raw <- files[["patientClinicalCourse"]] %>% filter(cohort == "AllAdm")

@@

```{r}
# Filtering on inpatients here (in_hospital == 1), to reduce subsequent compute time of the encounter length calculation 
clinical_raw <- filter(clinical_raw, in_hospital == 1)
```

@@

```{r}
icdCodes <- read.csv("public-data/pediatric_psychiatric_ICD10_codesV2.csv", header = TRUE, colClasses = "character") %>%
   filter( flag %in% c("originalCodes", "FrenchSpecificCodes")) %>%
   select( disorder_group, ICD10_Code, description )
```

### Calculate how many of the mental codes are in the dataset
mentalHealthCodes <- unique( obs_raw %>%
  dplyr::filter( concept_type == "DIAG-ICD10", 
                 concept_code %in% icdCodes$ICD10_Code ) %>%
  dplyr::mutate( concept_code = as.character( concept_code ), 
                 concept_length = nchar( concept_code ), 
                 concept_check = ifelse( concept_length == 4 | concept_length > 8, "Check", "OK")) %>%
  dplyr::select( concept_code, concept_length, concept_check) )

@@

obs_filtered <- obs_raw %>% 
  filter(concept_type == 'DIAG-ICD10', 
         days_since_admission >= 0)

clinical_filtered <- clinical_raw %>%
  filter( len_hospitalisation > min_hosp_days, 
          cohort  %in% c("AllAdm")) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic", "during_pandemic"))

# Steps necessary to filter:
  # - encounters without ICD
  # - ICD diagnosed outside of encounters
  # - patients from the patients table that don't have any encounters matching the criteria
clinical_table <- clinical_filtered %>% 
  inner_join( distinct(obs_filtered, patient_num, days_since_admission),  by = c("patient_num", "days_since_admission")) %>% 
  distinct(patient_num, n_hospitalisation) %>% 
  inner_join(clinical_filtered, by = c("patient_num", "n_hospitalisation"))
observation_table <- obs_filtered %>% 
  inner_join(distinct(clinical_table, patient_num, days_since_admission), by = c("patient_num", "days_since_admission"))
demographic_table <- demo_raw %>% 
  inner_join(distinct(clinical_table, patient_num), by = "patient_num")
all_codes <- observation_table %>% 
  inner_join(clinical_table, on = c("patient_num", "days_since_admission")) %>% 
  inner_join(select(demographic_table, -days_since_admission), on = c("patient_num"))
stopifnot(nrow(all_codes) == nrow(observation_table))
stopifnot(unique(all_codes$cohort) == "AllAdm")
stopifnot(unique(all_codes$in_hospital) == 1)
stopifnot(unique(all_codes$len_hospitalisation > min_hosp_days))

@@

### New hospitalizations
```{r, fig.width=12, fig.height=12}

new_hospitalizations_per_day <- clinical_filtered %>%
  group_by(patient_num, n_hospitalisation) %>%
  slice(which.min(date)) %>%
  filter( in_hospital == 1, 
         date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date, weeks, month) %>%
  summarise(count = n_distinct(patient_num)) %>% 
  ungroup()

@@

admission_plot_count <- function(df,
                               x,
                               y, 
                               title = "New admission counts",
                               color = NULL,
                               size = NULL) {
  plot <- ggplot(df, aes_string(x = x, y = y, color = color, size = size))  +
    geom_point() +
    scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
    theme(axis.text.x=element_text(angle=60, hjust=1))+
    labs(x = "Calendar date",
         y = "Patients count", 
       title = title)
  return(plot)
}

nh_day <- admission_plot_count(df = new_hospitalizations_per_day, 
                     x = "date",
                     y = "count", 
                     title = "New admissions (per day)")

nh_week <- admission_plot_count(df = new_hospitalizations_per_week, 
                     x = "weeks",
                     y = "count", 
                     title = "New admissions (per week)")

nh_month <- admission_plot_count(df = new_hospitalizations_per_month, 
                     x = "month",
                     y = "count", 
                     title = "New admissions (per month)")

grid.arrange(nh_day,
             nh_week,
             nh_month,
             ncol=1)
```

### Total hospitalize patients
```{r, fig.height=12, fig.width=12}
total_hospitalizations_per_day <- clinical_filtered %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count = n_distinct(patient_num))

unique_patients_hospitalized_per_week <- clinical_filtered %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks ) %>%
  summarise(count = n_distinct(patient_num))

unique_patients_hospitalized_per_month <- clinical_filtered %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month ) %>%
  summarise(count = n_distinct(patient_num))

@@

th_week <- ggplot(unique_patients_hospitalized_per_week, aes(x = weeks,
                       y = count))  +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(x = "Calendar date",
       y = "Patients count", 
       title = "Total admissions (per week)")

th_month <- ggplot(unique_patients_hospitalized_per_month, aes(x = month,
                       y = count))  +
  geom_line() +
  scale_x_date(date_breaks = "1 month", date_labels =  "%b %Y") +
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  labs(x = "Calendar date",
       y = "Patients count", 
       title = "Currently hospitalized patients (per month)")

@@

### Add the psychiatric/mental health ICD code description and filter 
```{r message=FALSE, warning=FALSE}
# Getting site name, will be used later in the analysis
site <- unique(as.character(demo_raw$siteid))

@@

all_codes_desc <- left_join( all_codes_desc, icdCategory) %>%
  mutate( icd_description_category = replace_na( ifelse( concept_code %in% icdCodes$ICD10_Code, "Psy", icd_description_category), "Others")) 
all_codes_psy <- all_codes_desc %>% 
  filter(concept_code %in% icdCodes$ICD10_Code)

if( raceAvailable == TRUE ){
  all_codes_psy <- all_codes_psy %>%
    left_join( race_raw, by = c("patient_num", "cohort") )
}
```

@@

```{r}
count_icd_per_day_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count_psy = n_distinct(patient_num)) %>%
  mutate( period = ifelse( date < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

count_icd_per_day <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count_icd = n_distinct(patient_num)) %>%
  mutate( period = ifelse( date < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

count_icd_per_day <- count_icd_per_day %>%
  left_join(count_icd_per_day_psy, 
            by = c("date", "period")) %>% 
  replace_na(list(count_psy = 0))

count_icd_per_day$percentage_psy <- count_icd_per_day$count_psy / count_icd_per_day$count_icd

thpd_psyPer <- count_icd_per_day %>%
  ggplot(aes(x = date, y = percentage_psy, fill = period)) +
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  labs(y = "Percentage",
       x = "Date (by day)",
       title = "Percentage of patients with mental health related ICD codes (daily)") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyPer
```

```{r}
thpd_psyCount <- count_icd_per_day %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count_icd)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = date, y = value, fill = counts, color = counts)) +
  geom_point() +
  # geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Date (by day)",
       title = "Daily patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount
```
